On the Construction of Isomorphic Interpreters (Journal Version)

### 2013-03-10 rc2012/iso-int.tex
reorg

aka. ultra-constructivism.

>>>>

(Maybe this section should be a seperate paper altogether)

This section introduces another class of isomorphic interpreters
{{langRevEE}} that arise from algebraic manipulations. Recursive type
definitions are viewed as algebraic polynomials characterizing the
shape of the type. Thus binary trees {{@@x.1+x*x}} are represented as
the polynomial {{x=1+x^2}} and hence {{x^2-x+1=0}}.

TODO: find a few more algebraic manipulations that give useful
combinators. Depending on which language we are working in, there may
be many. Suggestions:

{{ a * b <-> (-a) * (-b)}}

{{ -(-a) <-> a }}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Deriving the {{just : a <-> a + 1}} combinator}

In the spirit of the above discussion, the type {{nat}} defined by
{{@@x.1+x}} induces the polynomial {{nat = 1 + nat}}. Consider the
following manipulation:

{{a = a}}

{{a + nat = a + nat}}

{{a + nat = a + nat + 1}} because we know that {{nat = nat + 1}}.

{{a = a + 1}} by cancelling nat.

\noindent
These algebraic manipulations can be treated as a computation in
{{langRevT}} by witnessing each transformation with an appropriate
combinator.

\begin{itemize}
\item Thus we start with the the line corresponding to {{a=a}}:

DIAGRAM

\item And then we construct {{a * nat = a * nat}} by parallel
  composing with {{nat = nat}}.

DIAGRAM

\item And then we {{unfold}} on the RHS

DIAGRAM

\item And finally we use {{trace}} to do the appropriate cancellation,
  thereby getting the final circcuit, which we call {{just}}:

DIAGRAM

\end{itemize}

\noindent
Computationally the {{langRevT}} circuit does the following:

\begin{enumerate}
\item {{forall (v:a).just v |--> right v}} 
\item {{forall (v:a).just^{dagger} (right v) |--> v}} 
\item {{just (left ())}} diverges. 
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Isomorphisms of Trees}

Algebraic manipulations like the above have questionable mathematical
validitiy induced by the fact that the polynomial {{nat = nat + 1}} is
not well defined. Hence axiomatically adding such a polynomial may be
suspected as the underlying reason for the divergence.

In this context let us examine the isomorphism from one binary tree to
seven binary trees that is examined in the Fiore's paper
\cite{Fiore:2004}. Binary trees imply the polynomial {{t^2-t+1=0}}
which does have a valid root, namely {{t = 1/2 + i*sqrt(3)/2}}.

ALGEBRA

etc.

Which gives us the circuit:

DIAGRAM

Indeed this circuit is proof of the embedding of trees in a strong
constructive sense i.e. we can supply a binary tree and examine the
resulting seven binary trees. Further the adjoint of the circuit
allows us to supply seven binary trees and get one resultant tree.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Non-constructive Isomorphism of Trees}
However not all valid algebraic manipulations of a binary trees also
have meanignful computational interpretations. Consider the same
isomophism explained with the algebraic manipulation:

{{t^6 = 1}}

{{t^7 = t}}

The construction underlying this isomorphism is {{t^3 = -1}}, which
can be expressed in {{langRevEE}} as the cicruit:
